---
title: Drug Usage Exploration
format: closeread-html
fontsize: 2em
execute: 
    fig-width: 15
    fig-height: 12
---



<style>
body {
  background-color: #FAF3E0 ;
}
.cr-section {
  background-color: #FFE5B4 ;
}
</style>


# Drug use is not evenly distributed across the population. It varies first by broad demographics, then by education, and most strongly by personality/psychological traits like sensation seeking, impulsivity, and openness.. 

# Section 1: Introduction/Dataset Context 


```{r}
# load libraries 
library(tidyverse)
library(ggplot2)
library(scales)
library(gt)


# Load dataset
load("data.rda")

# See objects
ls()

# Assign correct dataset
drug_data <- da36536.0001

# Inspect
glimpse(drug_data)
summary(drug_data)


# Create any drug use var to represent users of any drug in the dataset
drug_data <- drug_data |>
  mutate(
    any_drug_use = ifelse(
      CANNABIS != "CL0" | COKE != "CL0" | ECSTASY != "CL0",
      "User", "Non-user"
    )
  )

  # create a small overview table describing the dataset
  overview_tbl <- tibble(
    Category = c(
      "Number of respondents",
      "Number of variables",
      "Drug use definition",
      "Drug variables used for current prototype",
      "Demographic variables",
      "Personality variables",
      "Survey population"
  ),
    Description = c(
      nrow(drug_data),
      ncol(drug_data),
      "User = respondent reported use of cannabis, coke, or ecstasy",
      "Cannabis, coke, ecstasy",
      "Age, gender, education, country, ethnicity",
      "Neuroticism, extraversion, openness, agreeableness, conscientiousness, impulsivity, sensation seeking",
      "Adults from English-speaking countries"
    )
)

# create formatted table using gt package
overview_tbl |>
  gt() |>
  tab_header(
    title = "Dataset Overview",
    subtitle = "Drug Consumption Survey Dataset"
  ) |>
  cols_width(
    Category ~ px(280),
    Description ~ px(520)
  )
```

This dataset contains individual survey responses about demographics, personality traits, and drug consumption. For this prototype, we define a respondent as a “User” if they reported any use of cannabis, coke, or ecstasy. This gives us a simplified outcome variable that lets us compare drug use patterns across demographic and psychological factors. 

<!-- # Section 2: Baseline Drug Use Patterns 
Communication Goal -> Establish the basic question: who reports drug use at all? -->

```{r}
  # calculate counts and percentages of users vs non-users
overall_usage <- drug_data |>
  # count number of respondents in each category
  count(any_drug_use) |>
  # create percentage column
  mutate(
    percent = n / sum(n)
  )

# create bar chart
overall_plot <- overall_usage |>
  ggplot(
    aes(
      x = any_drug_use,
      y = percent,
      fill = any_drug_use
    )) +
  geom_col(width = 0.6) +
  # add percentage labels on top of bars
  geom_text(
    aes(label = percent(percent, accuracy = 0.1)),
    vjust = -0.5,
    size = 5) +
  scale_y_continuous(
    labels = percent,
    limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70"
    )) +
  labs(
    title = "Overall Reported Drug Use",
    subtitle = "Users are respondents who reported cannabis, coke, or ecstasy use",
    x = "",
    y = "Percent of respondents",
    fill = "Drug usage") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )

```


```{r}
# Create summary table for age and drug use
age_summary <- drug_data |>
  mutate(
    age_group = case_when(
      AGE == "Ag1" ~ "18-24",
      AGE == "Ag2" ~ "25-34",
      AGE == "Ag3" ~ "35-44",
      AGE == "Ag4" ~ "45-54",
      AGE == "Ag5" ~ "55-64",
      AGE == "Ag6" ~ "65+",
      TRUE ~ NA_character_
    ),
    # keep age groups in the correct order
    age_group = factor(
      age_group,
      levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
    )
  ) |>
  # count users and non-users within each age group
  count(age_group, any_drug_use) |>
  # calculate percentage within each age group
  group_by(age_group) |>
  mutate(
    percent = n / sum(n)
  ) |>
  ungroup()

# Create plot 
age_plot <- ggplot(
  age_summary,
  aes(
    x = age_group,
    y = percent,
    fill = any_drug_use
  )) +
  geom_col(width = 0.75) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70"
    )) +
  labs(
    title = "Drug Use Is Highest Among Younger Respondents",
    subtitle = "Share of users and non-users within each age group",
    x = "Age range",
    y = "Percent of respondents",
    fill = "Drug use") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```


```{r}
# Create summary table for gender and drug use
gender_summary <- drug_data |>
  mutate(
    gender = case_when(
      GENDER == "M" ~ "Male",
      GENDER == "F" ~ "Female",
      TRUE ~ as.character(GENDER)
    )
  ) |>
  count(gender, any_drug_use) |>
  group_by(gender) |>
  mutate(
    percent = n / sum(n)
  ) |>
  ungroup()
  
# plot
gender_plot <- ggplot(
  gender_summary,
  aes(
    x = gender,
    y = percent,
    fill = any_drug_use)) +
  geom_col(width = 0.65) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70")) +
  labs(
    title = "Reported Drug Use Differs by Gender",
    subtitle = "Share of users and non-users within each gender group",
    x = "Gender",
    y = "Percent of respondents",
    fill = "Drug use") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right")

```


```{r}
# Create summary table for age, gender, and drug use
age_gender_summary <- drug_data |>
  mutate(
    age_group = case_when(
      AGE == "Ag1" ~ "18-24",
      AGE == "Ag2" ~ "25-34",
      AGE == "Ag3" ~ "35-44",
      AGE == "Ag4" ~ "45-54",
      AGE == "Ag5" ~ "55-64",
      AGE == "Ag6" ~ "65+"),
    age_group = factor(
      age_group,
      levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")),
    gender = case_when(
      GENDER == "M" ~ "Male",
      GENDER == "F" ~ "Female")
  ) |>
  count(gender, age_group, any_drug_use) |>
  group_by(gender, age_group) |>
  mutate(
    percent = n / sum(n)
  ) |>
  ungroup()

# Create age-by-gender plot
agebygender_plot <- ggplot(
  age_gender_summary,
  aes(x = age_group, y = percent, fill = any_drug_use)) +
  geom_col(width = 0.75) +
  facet_wrap(~ gender) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70")) +
  labs(
    title = "Age Patterns Differ Across Gender",
    subtitle = "Share of users and non-users within each age and gender group",
    x = "Age range",
    y = "Percent of respondents",
    fill = "Drug use") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right")

```


<!-- END OF section 1 & 2 ---------------------------------------------------------------------------    -->


<!-- section 4 -->

```{r}
drug_total <- drug_data |>
  pivot_longer(
    cols = c(ALCOHOL, AMPHET, AMYL, BENZOS, CAFF, CANNABIS, CHOC, 
             COKE, CRACK, ECSTASY, HEROIN, KETAMINE, LEGALH, LSD, 
             METH, MUSHROOMS, NICOTINE, SEMER, VSA), 
    names_to = "name",
    values_to = "usage"
  ) |>
  mutate(any_use = ifelse(usage == "CL0", 0,1))

# creating the graph
drug_total_plot <- drug_total |>
  group_by(name) |>
  summarize(percent_user = mean(any_use == 1)) |>
  ggplot(aes(x = reorder(name, percent_user), y = percent_user)) +
  geom_col(fill = "#F07A70") +
  coord_flip() +
  scale_y_continuous(labels = percent) +
  labs(
    title = "Most Commonly Reported Drug Usage",
    x = "Drug",
    y = "Percent of Users Reporting Use", 
    subtitle = "Share of Users for Each Drug"
  ) +
  theme_minimal(base_size=14)
```

```{r}

drug_freq <- drug_data |>
  pivot_longer(
    cols = c(ALCOHOL, AMPHET, AMYL, BENZOS, CAFF, CANNABIS, CHOC, COKE, CRACK, ECSTASY, HEROIN, KETAMINE, LEGALH, LSD, 
    METH, MUSHROOMS, NICOTINE, SEMER, VSA), 
    names_to = "name",
    values_to = "freq"
  ) |>
  count(name, freq) |>
  group_by(name) |>
  mutate(percent = n / sum(n)) |>
  ungroup()


#creating graph var
drug_freq_plot <- ggplot(drug_freq, aes(x = freq, y = name, fill = percent)) +
  geom_tile() +
  scale_fill_gradient(low = "#FFF5F7", high = "#F07A70") +
  scale_x_discrete(labels = c(
    "CL0" = "Never Used",
    "CL1" = "Used 10+ Years Ago", 
    "CL2" = "Used in Last Decade", 
    "CL3" = "Used in Last Year", 
    "CL4" = "Used in Last Month", 
    "CL5" = "Used in Last Week" , 
    "CL6" = "Used in Last Day")) +
  labs(
    title = "Drug Use Frequency Across Substances",
    x = "Reported Frequency Category",
    y = "Drug",
    fill = "Percent"
  )
```

<!-- end of section 4 -->


::: {.cr-section}

:::{.narrative}
## Section 2: Baseline Drug Use Patterns
## Lets establish the basic question: who reports drug use at all?
:::


This graph establishes the baseline distribution of drug usage within the dataset using our simplified classification system. Respondents were categorized as “Users” if they reported any cannabis, coke, or ecstasy use, while all others were labeled “Non-users.” Almost 80% of respondents fall into the “User” category, suggesting that some level of reported drug use is relatively common within this survey population. This baseline helps contextualize the demographic and personality-based patterns explored in later sections of the project. @cr-overall

:::{#cr-overall}
```{r}
overall_plot
```
:::

This graph shows a strong relationship between age and reported drug use within the survey population. Respondents in the youngest age group, 18–24, had the highest share of reported drug use, with usage steadily decreasing across older age groups. By the 65+ category, non-users make up the majority of respondents. This pattern suggests that younger individuals in the dataset are substantially more likely to report drug use than older respondents, making age one of the clearest demographic trends in the project. @cr-age


::: {#cr-age}
```{r}
age_plot
```
:::

This graph shows noticeable differences in reported drug use across gender groups within the dataset. Male respondents report higher levels of drug use overall, with almost 90% categorized as “Users,” compared to a lower share among female respondents. While both groups contain a majority of users, the gap suggests that gender may play a role in reported drug consumption patterns. Together with the age analysis, this graph helps establish demographic trends before exploring more detailed personality-based relationships later in the project. @cr-gender

::: {#cr-gender}
```{r}
gender_plot
```
:::

This graph combines the project’s two strongest demographic patterns: age and gender. Across both male and female respondents, reported drug use is highest among younger age groups and generally declines with age. However, the decline is more pronounced among female respondents, where the proportion of non-users increases substantially in older age categories. Male respondents consistently report higher levels of drug use across nearly every age group. By comparing both demographics simultaneously, this visualization shows that age-related patterns remain visible across genders while also highlighting an overall gender gap in reported drug use. @cr-agegender


::: {#cr-agegender}
```{r}
agebygender_plot
```
:::

:::{.narrative}
## Section 3: Education and Drug Use
## Does social/demographic structure beyond age and gender relate to drug use?
:::
---- ADD graphs + analysis here (follow earlier structure) ---

:::{.narrative}
## Section 4: Which Drugs are Most Common?
## Let's examine which drugs are most commonly reported.
:::


Drug use is a broad term. This graph establishes that the most commonly used drugs across respondents is caffeine, chocolate, and alchol, with almost an 100% rate. As such, this data demonstrates that most people who answer the survey will be categorized as a drug user. Over 75% of users reported consuming cannabis and nicotine and many of the illegal drugs reported a lower percentage of users. These patterns suggest that the most commonly used substances are typically legal and easy to obtain. It also demonstrates a social aspect to drug usage as the majority of the most commonly used drugs are socially acceptable to consume. @cr-drugtotal

::: {#cr-drugtotal}
```{r}
drug_total_plot
```
:::

This graph shows that most drugs are not used frequently. However, if respondents are using certain drugs with a frequency, those drugs tend to be alcohol, caffeine, cannabis, chocolate, and nicotine. As shown with the drug use graph, this data helps to reinforce that socially acceptable, legal drugs are utilized more. This graph also helps to show a difference between reporting use of a drug and actually consuming it regularly. While an individual may report consuming a drug, it may be experimental (once or twice) rather than frequent. Some drugs may also have less people using them, but they are using them with more frequency - like mushrooms or ecstasy. @cr-drugfrequency

::: {#cr-drugfrequency}
```{r}
drug_freq_plot
```
:::

:::{.narrative}
## Section 5: Personality Traits & Drug Use
:::
---- ADD graphs + analysis here (follow earlier structure) ---

:::{.narrative}
## Section 6: Personality Traits by Specific Drugs
:::
---- ADD graphs + analysis here (follow earlier structure) ---

:::{.narrative}
## Section 7: Conclusion
:::


:::