---
title: Drug Usage Exploration
format: closeread-html
fontsize: 2em
execute: 
    fig-width: 15
    fig-height: 12
---


<style>
body {
  background-color: #FAF3E0 ;
}
.cr-section {
  background-color: #FFE5B4 ;
}
</style>

# Section 1: Introduction/Dataset Context
 Drug use is not evenly distributed across the population. It varies first by broad demographics, then by education, and most strongly by personality/psychological traits like sensation seeking, impulsivity, and openness.

```{r}
# load libraries 
library(tidyverse)
library(ggplot2)
library(scales)
library(closeread)
library(gt)


# Load dataset
load("data.rda")

# See objects
ls()

# Assign correct dataset
drug_data <- da36536.0001

# Inspect
glimpse(drug_data)
summary(drug_data)


# Create any drug use var to represent users of any drug in the dataset
drug_data <- drug_data |>
  mutate(
    any_drug_use = ifelse(
      CANNABIS != "CL0" | COKE != "CL0" | ECSTASY != "CL0",
      "User", "Non-user"
    )
  )

  # create a small overview table describing the dataset
  overview_tbl <- tibble(
    Category = c(
      "Number of respondents",
      "Number of variables",
      "Drug use definition",
      "Drug variables used for current prototype",
      "Demographic variables",
      "Personality variables",
      "Survey population"
  ),
    Description = c(
      nrow(drug_data),
      ncol(drug_data),
      "User = respondent reported use of cannabis, coke, or ecstasy",
      "Cannabis, coke, ecstasy",
      "Age, gender, education, country, ethnicity",
      "Neuroticism, extraversion, openness, agreeableness, conscientiousness, impulsivity, sensation seeking",
      "Adults from English-speaking countries"
    )
)

# create formatted table using gt package
overview_tbl |>
  gt() |>
  tab_header(
    title = "Dataset Overview",
    subtitle = "Drug Consumption Survey Dataset"
  ) |>
  cols_width(
    Category ~ px(280),
    Description ~ px(520)
  )
```

This dataset contains individual survey responses about demographics, personality traits, and drug consumption. For this prototype, we define a respondent as a “User” if they reported any use of cannabis, coke, or ecstasy. This gives us a simplified outcome variable that lets us compare drug use patterns across demographic and psychological factors.

# Section 2: Baseline Drug Use Patterns
Communication Goal -> Establish the basic question: who reports drug use at all?

```{r}
  # calculate counts and percentages of users vs non-users
overall_usage <- drug_data |>
  # count number of respondents in each category
  count(any_drug_use) |>
  # create percentage column
  mutate(
    percent = n / sum(n)
  )

# create bar chart
overall_usage |>
  ggplot(
    aes(
      x = any_drug_use,
      y = percent,
      fill = any_drug_use
    )) +
  geom_col(width = 0.6) +
  # add percentage labels on top of bars
  geom_text(
    aes(label = percent(percent, accuracy = 0.1)),
    vjust = -0.5,
    size = 5) +
  scale_y_continuous(
    labels = percent,
    limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70"
    )) +
  labs(
    title = "Overall Reported Drug Use",
    subtitle = "Users are respondents who reported cannabis, coke, or ecstasy use",
    x = "",
    y = "Percent of respondents",
    fill = "Drug usage") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )

```

This graph establishes the baseline distribution of drug usage within the dataset using our simplified classification system. Respondents were categorized as “Users” if they reported any cannabis, coke, or ecstasy use, while all others were labeled “Non-users.” Almost 80% of respondents fall into the “User” category, suggesting that some level of reported drug use is relatively common within this survey population. This baseline helps contextualize the demographic and personality-based patterns explored in later sections of the project.

```{r}
# Create summary table for age and drug use
age_summary <- drug_data |>
  mutate(
    age_group = case_when(
      AGE == "Ag1" ~ "18-24",
      AGE == "Ag2" ~ "25-34",
      AGE == "Ag3" ~ "35-44",
      AGE == "Ag4" ~ "45-54",
      AGE == "Ag5" ~ "55-64",
      AGE == "Ag6" ~ "65+",
      TRUE ~ NA_character_
    ),
    # keep age groups in the correct order
    age_group = factor(
      age_group,
      levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")
    )
  ) |>
  # count users and non-users within each age group
  count(age_group, any_drug_use) |>
  # calculate percentage within each age group
  group_by(age_group) |>
  mutate(
    percent = n / sum(n)
  ) |>
  ungroup()

# Create plot 
ggplot(
  age_summary,
  aes(
    x = age_group,
    y = percent,
    fill = any_drug_use
  )) +
  geom_col(width = 0.75) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70"
    )) +
  labs(
    title = "Drug Use Is Highest Among Younger Respondents",
    subtitle = "Share of users and non-users within each age group",
    x = "Age range",
    y = "Percent of respondents",
    fill = "Drug use") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```

This graph shows a strong relationship between age and reported drug use within the survey population. Respondents in the youngest age group, 18–24, had the highest share of reported drug use, with usage steadily decreasing across older age groups. By the 65+ category, non-users make up the majority of respondents. This pattern suggests that younger individuals in the dataset are substantially more likely to report drug use than older respondents, making age one of the clearest demographic trends in the project.

```{r}
# Create summary table for gender and drug use
gender_summary <- drug_data |>
  mutate(
    gender = case_when(
      GENDER == "M" ~ "Male",
      GENDER == "F" ~ "Female",
      TRUE ~ as.character(GENDER)
    )
  ) |>
  count(gender, any_drug_use) |>
  group_by(gender) |>
  mutate(
    percent = n / sum(n)
  ) |>
  ungroup()
  
# plot
ggplot(
  gender_summary,
  aes(
    x = gender,
    y = percent,
    fill = any_drug_use)) +
  geom_col(width = 0.65) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70")) +
  labs(
    title = "Reported Drug Use Differs by Gender",
    subtitle = "Share of users and non-users within each gender group",
    x = "Gender",
    y = "Percent of respondents",
    fill = "Drug use") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right")

```
This graph shows noticeable differences in reported drug use across gender groups within the dataset. Male respondents report higher levels of drug use overall, with almost 90% categorized as “Users,” compared to a lower share among female respondents. While both groups contain a majority of users, the gap suggests that gender may play a role in reported drug consumption patterns. Together with the age analysis, this graph helps establish demographic trends before exploring more detailed personality-based relationships later in the project.

```{r}
# Create summary table for age, gender, and drug use
age_gender_summary <- drug_data |>
  mutate(
    age_group = case_when(
      AGE == "Ag1" ~ "18-24",
      AGE == "Ag2" ~ "25-34",
      AGE == "Ag3" ~ "35-44",
      AGE == "Ag4" ~ "45-54",
      AGE == "Ag5" ~ "55-64",
      AGE == "Ag6" ~ "65+"),
    age_group = factor(
      age_group,
      levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")),
    gender = case_when(
      GENDER == "M" ~ "Male",
      GENDER == "F" ~ "Female")
  ) |>
  count(gender, age_group, any_drug_use) |>
  group_by(gender, age_group) |>
  mutate(
    percent = n / sum(n)
  ) |>
  ungroup()

# Create age-by-gender plot
ggplot(
  age_gender_summary,
  aes(x = age_group, y = percent, fill = any_drug_use)) +
  geom_col(width = 0.75) +
  facet_wrap(~ gender) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "User" = "#20B8C4",
      "Non-user" = "#F07A70")) +
  labs(
    title = "Age Patterns Differ Across Gender",
    subtitle = "Share of users and non-users within each age and gender group",
    x = "Age range",
    y = "Percent of respondents",
    fill = "Drug use") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right")

```
This graph combines the project’s two strongest demographic patterns: age and gender. Across both male and female respondents, reported drug use is highest among younger age groups and generally declines with age. However, the decline is more pronounced among female respondents, where the proportion of non-users increases substantially in older age categories. Male respondents consistently report higher levels of drug use across nearly every age group. By comparing both demographics simultaneously, this visualization shows that age-related patterns remain visible across genders while also highlighting an overall gender gap in reported drug use.


OUTLINE ENDS HERE ---------------------------------------------------------------------------

```{r}
# Create plot examining education and drug use
plot_2 <- ggplot(drug_data, aes(x = EDUCATION, fill = any_drug_use)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Overall Drug Usage by Education Level",
    x = "Education Level",
    y = "Percent",
    fill = "Drug Usage"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
# Create plot examining sensation seeking and selected drug use
plot_4 <- ggplot(drug_data, aes(x = SS, fill = any_drug_use)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Drug Usage by Sensation Seeking Score",
    x = "Sensation Seeking Score",
    y = "Percent",
    fill = "Drug Usage"
  ) +
  theme_minimal()
```



::: {.cr-section}

:::{.narrative}
Drug usage varies across age, gender, education, and personality traits. We aim to study these patterns by analyzing data from a year-long survey which collected the drug consumption habits of individuals 18 years or older. The survey questioned if respondents had consumed alcohol, amphetamines, amyl nitrite, benzodiazepine, cannabis, chocolate, cocaine, caffeine, crack, ecstasy, heroin, ketamine, legal highs, LSD, methadone, mushrooms, nicotine, and volatile substance abuse and one fictional drug (Semeron). 

:::

Within the study participants, a higher percentage of the male respondents use drugs versus the female interviewees. @cr-genderplot

::: {#cr-genderplot}
```{r}
plot_1
```
:::

Within the recorded age ranges, the higest percentage of indviduals who utilize drugs fall in the 18-24 age range. The percentage of respondents who utilize drugs decreases with each age range, with 65+ being the lowest. @cr-ageplot

::: {#cr-ageplot}
```{r}
plot_3
```
:::

 The highest percentage of drug users falls in the category of attending some university or college but not obtaining a degree. Once an individual has gained a bachelors degree or higher, the percentage of drug users in these categories tend to stay around 70%. The category of those who left school at 16 years of age (Ed2) also has a similar percentage of drug users as the groups which earned a bachelors or higher. This suggests that the level of education an individual achieves can affect drug consumption, but it is not the only factor. @cr-educationplot


::: {#cr-educationplot}
```{r}
plot_2
```
:::

Add analysis here @cr-personalityplot

::: {#cr-personalityplot}
```{r}
plot_4
```
:::

:::